pyTopoComplexity: A Python package for topographic complexity analysis

### 2024-08-29 JOSS/paper.rmd
Add files via upload

---
title: 'pyTopoComplexity: A Python package for topographic complexity analysis'
tags:
  - Python
  - topographic complexity
  - surface roughness
  - two-dimensional continuous wavelet transform
  - fractal dimension
  - Rugosity Index
  - Terrain Position Index
authors:
  - name: Larry Syu-Heng Lai
    corresponding: true
    orcid: 0000-0003-4589-9729 
    affiliation: "1"
  - name: Adam M. Booth
    orcid: 0000-0002-7339-0594   
    affiliation: "2"
  - name: Alison R. Duvall
    orcid: 0000-0002-7760-7236 
    affiliation: "1"
  - name: Erich Herzig
    orcid: 0000-0002-3898-5971 
    affiliation: "1"
affiliations:
 - name: Department of Earth and Space Sciences, University of Washington, Seattle, Washington, USA
   index: 1
 - name: Department of Geology, Portland State University, Portland, Oregon, USA
   index: 2
date: 18 August 2024
bibliography: paper.bib
---

@@

**pyTopoComplexity** is a Python package that provides a computationally efficient and customizable 
implementation of four modern methods for quantifying topographic complexity. These methods include 
two-dimensional continuous wavelet transform (2D-CWT) analysis, fractal dimension estimation, Rugosity 
Index and Terrain Position Index calculations across various spatial scales. This package addresses 
the scarcity of open-source software for these sophisticated methods, which are crucial in modern 
terrain analysis, and facilitates data comparison and reproducibility. In the [software respository](https://github.com/GeoLarryLai/pyTopoComplexity.git), 
we also include a Jupyter Notebook file that integrates components from the Python-based 
surface-process modeling platform **Landlab** [@Hobley2017]. This allows researchers to 
simulate the smoothing of topography over time through terrestrial nonlinear hillslope 
diffusion processes. By combining these features, **pyTopoComplexity** advances the toolset 
available to researchers for measuring and simulating the time-dependent persistence of 
topographic complexity signatures against environmental forces on terrain surfaces.

@@

Topographic complexity, often referred to as topographic roughness or surface roughness, provides 
critical insights into surface processes and the interactions among the geosphere, biosphere, and 
hydrosphere. With the increasing availability, utility, and popularity of digital terrain model (DTM) 
data, quantifying topographic complexity has become an essential measure in terrain analysis across 
various research fields. This necessity spans applications such as terrain classification and mapping 
at various spatial scales [@Weiss2001; @Robbins2018; @Lindsay2019; @PardoIguzquiza2022a], evaluating 
the depositional age of event sedimentation and subsequent erosion processes 
[@Hetz2016; @Johnstone2018; @Booth2017; @LaHusen2020; @Herzig2023], and identifying habitats to assess 
ecological diversity on land and seafloor [@Frost2005; @Hetz2016; @Wilson2007].

@@

To address this gap, we have developed an open-source Python toolkit called **pyTopoComplexity**. This toolkit 
offers computationally efficient and easily customizable implementations for performing and visualizing the results 
of 2D-CWT, fractal dimension, and rugosity index calculations (see Table 1\autoref{tab:1}). Additionally, **pyTopoComplexity** 
includes a module for calculating the Terrain Position Index (TPI), a widely used metric that is considered effective 
in geomorphology research [@Newman2018; @Deumlich2010; @Liu2011] and often used alongside Rugosity Index in marine 
geological and ecological studies [@Wilson2007; @Walbridge2018].

This toolkit can detect the grid spacing and unit of the projected coordinate system (acceptable in meters, 
U.S. survey feet, and international feet) from the input raster DTM file (GeoTIFF format) and 
automatically conduct unit conversions in necessary calculation steps to ensure data consistency and 
reproducibility. Results at nodes affected by edge effects due to no-data values outside the input raster 
will be removed by default. Users can define the suitable spatial scale to match their research purposes 
and choose computational approaches (e.g., chunk processing, faster mathematical approximations) to 
optimize performance (see details in the **Methods and features overview** section).

| Modules  | Classes | Method Descriptions | References |
| ------------- | ------------- | ------------- | ------------- |
| pycwtmexhat.py | CWTMexHat | Quantifies the wavelet-based curvature of the land surface using two-dimensional continuous wavelet transform (2D-CWT) with a Mexican Hat wavelet | @Booth2009; @Booth2017 |
| pyfracd.py | FracD | Conducts fractal dimension analysis on the land surface using variogram procedures | @Wen1997; @PardoIguzquiza2020|
| pyrugostiy.py | RugosityIndex | Calculates the Rugosity Index of the land surface | @Jenness2004; @DuPreez2015 |
| pytpi.py | TPI | Calculate Terrain Position Index of the land or seafloor surface | @Weiss2001; @Walbridge2018 |
Table: Table 1 \label{tab:1}: Modules contained in the **pyTopoComplexity** package.

Each module of the **pyTopoComplexity** is provided with a corresponding [example Jupyter Notebook file](https://github.com/GeoLarryLai/pyTopoComplexity/tree/main/example) 
for usage instructions, using the Light Detection and Ranging (LiDAR) DTM data of a deep-seated landslide that 
occurred in 2014 in the Oso area of the North Fork Stillaguamish River valley, Washington State, USA 
[@WashingtonGeologicalSurvey2023]. In the software repository, we also include an additional Jupyter Notebook file 
[**nonlineardiff_Landlab.ipynb**](https://github.com/GeoLarryLai/pyTopoComplexity/blob/main/example/nonlineardiff_Landlab.ipynb), 
which allows researchers to simulate the smoothing of topography over time via terrestrial nonlinear hillslope 
diffusion processes [@Roering1999]. This is achieved by employing the `TaylorNonLinearDiffuser` module from the 
`terrainBento` Python package [@Barnhart2019] and running the simulation in the **Landlab** environment [@Hobley2017].

@@

The **pycwtmexhat.py** module in **pyTopoComplexity** implements the 2D-CWT method for terrain analysis, providing detailed information on how 
amplitude is distributed across spatial frequencies at each position in the data by transforming spatial data into position-frequency 
space. This method is particularly effective for depicting the Laplacian of topography [@Torrence1998; @Lashermes2007], revealing 
concave and convex regions of topography at various scales [@Malamud2001; @Struble2021], identifying deep-seated landslides 
[@Booth2009; @Berti2013], and estimating the depositional ages of landslide deposits [@Booth2017; @LaHusen2020; @Underwood2022; @Herzig2023].

@@

, where the resultant wavelet coefficient $C(s,x,y)$ provides a measure of how well the wavelet $\psi$ matches the data $z$ at each grid [@Torrence1998]. 
When $s$ is large, $\psi$ is spread out, capturing long-wavelength features of $z$; when $s$ is small, $\psi$ becomes more localized, making it sensitive 
to fine-scale features of $z$. In this implementation, we use the 2D Mexican Hat wavelet (i.e., Ricker wavelet) function to define $\psi$:

@@

The Mexican hat wavelet is proportional to the second derivative of a Gaussian envelope, with its Fourier wavelength ($\lambda$) 
dependent on the grid spacing ($\Delta$) of the input DTM raster. The wavelet function $\psi$ is scaled according to the 
wavelet scale parameter $s$ and the grid spacing $\Delta$, ensuring that the resultant wavelet coefficient $C$ signifies 
concave and convex landforms corresponding to the wavelet scale $s$. Users can define the value of $\lambda$ in meters as the 
targeted spatial scale for landform roughness analysis, and the **pycwtmexhat.py** module will automatically compute the wavelet 
scale $s$ based on the grid spacing ($\Delta$) of the input raster file (Figure 1)\autoref{fig:1}.

We note that the $C$ and $\psi$ equations presented here are the mathmetical approaches adapted in later publications 
[@LaHusen2020; @Underwood2022; @Herzig2023] and ongoing works [@Booth2023; @Lai2023; @Ozioko2023] of landslide mapping and 
age dating studies, with minor differences from the earlier similar research in @Booth2009 and @Booth2017 (original MATLAB codes 
available from [Booth's personal website](https://web.pdx.edu/~boothad/tools.html)). These differences in methatical approach 
can result $C$ values in different units and order of magnitude (e.g., $10^{-3}$ to $10^{-4}$ [m$^{-2}$] in @Booth2017 
and prior studies; $10^{-2}$ to $10^{-3}$ [m$^{-1}$] in @LaHusen2020 thereafter). Despite this discrepency, the $C$ values 
yielded from these two approaches are linearly scaled and interconvertible, and they both reflect identical spatiotemporal patterns 
of topographic complexity (i.e., surface roughness).

@@

In this module, we adapt the variogram method to estimate the local fractal dimension within a moving window centered at each cell of 
the DTM [@Taud2005; @PardoIguzquiza2020; @PardoIguzquiza2022a]. This approach simplifies the problem to estimating the fractal dimension 
of one-dimensional topographic profiles [@Dubuc1989] within the two-dimensional moving window. For a one-dimensional profile of length $L$, 
the variogram $\gamma_1(k)$ can be estimated at the $K$ lag distances ($K$ lag distances ($k = 1, \ldots, K$) by:

@@

The **pyrugosity.py** module in **pyTopoComplexity**  measures the Rugosity Index of the land surface, which is widely used to assess structural 
complexity of the topography. Such method has been applied in classifying seafloor types by marine geologists and geomorphologist, understanding small-scale 
hydrodynamics by oceanographers, and studying available habitats in the landscape by ecologists and coral biologists [@Lundblad2006; @Wilson2007].

@@

This module adapts the Triangulated Irregular Networks method from @Jenness2004 to approximate the contoured area as the sum of 
eight truncated-triangle areas. These triangles connect the central grids, four corner grids, and four grids at the middle points of the 
surrounding edges within the moving window. If no local slope correction is applied, the planimetric area is considered to be the horizontal 
planar area of the moving window, as described in @Jenness2004. Another approach considers slope correction where to the planimetric area is 
projected onto an plane of the local gradient [@DuPreez2015].

By definition, the Rugosity Index is as a minimum value of one (completely flate surface). Typical valuesrange of the conventional rugosity 
index (without slope correction) from one to three although larger values are possible in very steep terrains. The slope-corrected rugosity 
index, also called arc-chord ratio (ACR) Rugosity Index, could provide a better representation of local surface complexity (Figure 3)\autoref{fig:3}.

@@

The **pytpi.py** module in **pyTopoComplexity** calculates the Terrain Position Index (TPI) of the land surface. The TPI, also known as the 
Topographic Position Index in terrestrial studies [@Weiss2001], measures the relative topographic position of a point compared to its 
surrounding landforms. This metric highlights regions that are relatively higher or lower than their surroundings, which is useful for 
distinguishing landscape features such as hilltops, valleys, flat plains, and slopes. In oceanography, an equivalent metric is the 
Bathymetric Position Index (BPI), which applies the TPI algorithm to bathymetric data to evaluate seafloor complexity.

@@

In this module, the TPI value is computed at the central grid within a square-shaped moving window. Users can specify the size of the moving 
window (i.e., the number of grids along each edge) to assess the topographic position at different spatial scales. Positive TPI values 
indicate elevated features, negative values represent depressions, and values close to zero denote flat or gently sloping areas. The **pytpi.py** 
module also returns the absolute values of the TPI (Figure 4\autoref{fig:4}). This absolute TPI raster is suggested to be effective for 
quick assessments in mapping and relative dating of deep-seated landslide deposits using high-resolution DTM data [@Lai2023].

@@

## Forward simulation of landscape smoothing through nonlinear hillslope diffusion process

@@

, where \(\mathbf{q}_s\) represents the sediment flux at the surface. The sediment flux is further defined by a nonlinear flux law that is 
approximated using a Taylor series expansion [@Ganti2012]:

@@

Here, $\mathbf{S} = -\nabla z$ represents the downslope topographic gradient, and $S$ is its magnitude. The parameter $K$ is a diffusion-like 
transport coefficient with dimensions of length squared per time. The simulation also incorporates the critical slope gradient ($S_c$) to 
ensure numerical stability and prevent the numerical instability when $S = S_c$. $N$ denotes the number of terms in the Taylor 
expansion, while $i$ specifies the number of additional terms included. If $N = 0$, the expression simplifies to linear diffusion [@Culling1963]. 
The default is set to $N = 2$ that gives the behavior described in @Ganti2012 as an approximation of the nonlinear diffusion.

This notebook provides a comprehensive workflow that guides users through setting up, importing raster files, and running simulations. Since 
**Landlab** primarily handles DTM data in ESRI ASCII format, this notebook includes utility functions for converting raster files between GeoTIFF 
and ESRI ASCII formats. Users are required to specify the values for $S_c$, $K$, the length of each time step in years, and the final time to 
stop the simulation. The example included in the notebook uses LiDAR DTM data from the 2014 Oso Landslide [@WashingtonGeologicalSurvey2023], 
with parameter values provided in @Booth2017, in an attempt to reproduce the simulation results presented in that study (Figure 5\autoref{fig:5}).

@@

The development of **pyTopoComplexity** is part of a collaborative effort within the [Landslide subteam](https://cascadiacopeshub.org/team-1-landslides/) of the 
[Cascadia Coastlines and Peoples Hazards Research Hub (Cascadia CoPes Hub)](https://cascadiacopeshub.org), funded by the National Science Foundation Award ([#2103713](https://www.nsf.gov/awardsearch/showAward?AWD_ID=2103713&HistoricalAwards=false)). 
We extend special thanks to Dr. Eulogio Pardo-Igúzquiza, who generously shared his Fortran code for fractal dimension analysis used in his work 
[@PardoIguzquiza2022b], which  inspired the development of the **pyfracd.py** module.

>>>>

---
title: 'pyTopoComplexity: A Python package for topographic complexity analysis'
tags:
  - Python
  - topographic complexity
  - surface roughness
  - two-dimensional continuous wavelet transform
  - fractal dimension
  - Rugosity Index
  - Terrain Position Index
authors:
  - name: Larry Syu-Heng Lai
    corresponding: true
    orcid: 0000-0003-4589-9729 
    affiliation: "1"
  - name: Adam M. Booth
    orcid: 0000-0002-7339-0594   
    affiliation: "2"
  - name: Alison R. Duvall
    orcid: 0000-0002-7760-7236 
    affiliation: "1"
  - name: Erich Herzig
    orcid: 0000-0002-3898-5971 
    affiliation: "1"
affiliations:
 - name: Department of Earth and Space Sciences, University of Washington, Seattle, Washington, USA
   index: 1
 - name: Department of Geology, Portland State University, Portland, Oregon, USA
   index: 2
date: 30 August 2024
bibliography: paper.bib
---

@@

**pyTopoComplexity** is a Python package that provides a computationally efficient and customizable 
implementation of four modern methods for quantifying topographic complexity. These methods include 
two-dimensional continuous wavelet transform (2D-CWT) analysis, fractal dimension estimation, Rugosity 
Index and Terrain Position Index calculations across various spatial scales. This package addresses 
the scarcity of open-source software for these sophisticated methods needed in modern 
terrain analysis, and facilitates data comparison and reproducibility. In the [software respository](https://github.com/GeoLarryLai/pyTopoComplexity.git), 
we also include a Jupyter Notebook file that integrates components from the Python-based 
surface-process modeling platform **Landlab** [@Hobley2017]. The notebook allows researchers to 
simulate the smoothing of topography over time through terrestrial nonlinear hillslope 
diffusion processes. By combining these features, **pyTopoComplexity** advances the toolset 
available to researchers for measuring and simulating the time-dependent persistence of 
topographic complexity signatures against environmental forces on terrain surfaces.

@@

Topographic complexity, often referred to as topographic roughness or surface roughness, provides 
critical insights into surface processes and the interactions among the geosphere, biosphere, and 
hydrosphere [@Dietrich2006]. With the increasing availability, utility, and popularity of digital terrain model (DTM) 
data, quantifying topographic complexity has become an essential step in terrain analysis across 
various research fields. This necessity spans applications such as terrain classification and mapping 
at various spatial scales [@Weiss2001; @Robbins2018; @Lindsay2019; @PardoIguzquiza2022a], evaluating 
the depositional age of event sedimentation (e.g., landslides, alluvial fans) and subsequent erosion processes 
[@Hetz2016; @Johnstone2018; @Booth2017; @LaHusen2020; @Herzig2023], and identifying habitats to assess 
ecological diversity on land and seafloor [@Frost2005; @Hetz2016; @Wilson2007].

@@

To address this gap, we have developed an open-source Python toolkit called **pyTopoComplexity**. This toolkit 
offers computationally efficient and easily customizable implementations for performing and visualizing the results 
of 2D-CWT, fractal dimension, and rugosity index calculations (see Table 1\autoref{tab:1}). Additionally, **pyTopoComplexity** 
includes a module for calculating the Terrain Position Index (TPI), a widely used metric in geomorphology 
research [@Newman2018; @Deumlich2010; @Liu2011] and often used alongside Rugosity Index in marine 
geological and ecological studies [@Wilson2007; @Walbridge2018].

This toolkit detects the grid spacing and units of the projected coordinate system (acceptable in meters, 
U.S. survey feet, and international feet) from the input raster DTM file (GeoTIFF format) and 
automatically conducts unit conversions in necessary calculation steps to ensure data consistency and 
reproducibility. Removal of results at nodes affected by edge effects due to no-data values outside the input 
raster occurs by default. Users can define the suitable spatial scale to match their research purposes 
and choose computational approaches (e.g., chunk processing, faster mathematical approximations) to 
optimize performance (see details in the **Methods and features overview** section).

| Modules  | Classes | Method Descriptions | References |
| ------------- | ------------- | ------------- | ------------- |
| pycwtmexhat.py | CWTMexHat | Quantifies the wavelet-based curvature of the land surface using two-dimensional continuous wavelet transform (2D-CWT) with a Mexican Hat wavelet | @Booth2009; @Booth2017 |
| pyfracd.py | FracD | Conducts fractal dimension analysis on the land surface using variogram procedures | @Wen1997; @PardoIguzquiza2020|
| pyrugostiy.py | RugosityIndex | Calculates the Rugosity Index of the land surface | @Jenness2004; @DuPreez2015 |
| pytpi.py | TPI | Calculates Terrain Position Index of the land or seafloor surface | @Weiss2001; @Walbridge2018 |
Table: Table 1 \label{tab:1}: Modules contained in the **pyTopoComplexity** package.

Each module of **pyTopoComplexity** includes a corresponding [example Jupyter Notebook file](https://github.com/GeoLarryLai/pyTopoComplexity/tree/main/example)
with usage instructions. These examples utilize lidar DTM data (~1 m resolution) from the 2014 ‘Oso’ deep-seated landslide
along the North Fork Stillaguamish River valley in Washington State, USA [@WashingtonGeologicalSurvey2023]. In the software repository, we also include an additional Jupyter Notebook file 
[**nonlineardiff_Landlab.ipynb**](https://github.com/GeoLarryLai/pyTopoComplexity/blob/main/example/nonlineardiff_Landlab.ipynb), 
which allows researchers to simulate the smoothing of topography over time via terrestrial nonlinear hillslope 
diffusion processes [@Roering1999]. This is achieved by employing the `TaylorNonLinearDiffuser` module from the 
`terrainBento` Python package [@Barnhart2019] and running the simulation in the **Landlab** environment [@Hobley2017].

@@

The **pycwtmexhat.py** module in **pyTopoComplexity** implements the 2D-CWT method for terrain analysis, providing detailed information on how 
amplitude is distributed across spatial frequencies at each position in the data by transforming spatial data into position-frequency 
space. When used with the Gaussian family of wavelets, this method is particularly effective for depicting the Laplacian of topography 
[@Torrence1998; @Lashermes2007], revealing concave and convex regions of topography at various smoothing-length scales [@Malamud2001; @Struble2021], 
identifying deep-seated landslides [@Booth2009; @Berti2013], and estimating the depositional ages of landslide deposits [@Booth2017; @LaHusen2020; @Underwood2022; @Herzig2023].

The **pycwtmexhat.py** module in **pyTopoComplexity** implements the 2D-CWT method for terrain analysis, providing detailed information on how 
amplitude is distributed across spatial frequencies at each position in the data by transforming spatial data into position-frequency 
space. When used with the Gaussian family of wavelets, this method is particularly effective for depicting the Laplacian of topography 
[@Torrence1998; @Lashermes2007], revealing concave and convex regions of topography at various smoothing-length scales [@Malamud2001; @Struble2021], 
identifying deep-seated landslides [@Booth2009; @Berti2013], and estimating the depositional ages of landslide deposits [@Booth2017; @LaHusen2020; @Underwood2022; @Herzig2023]. 
Unlike techniques that use moving windows to detect regional topographic complexity, the 2D-CWT method stands out by isolating specific landform features at the scale of the 
designated $\lambda$ while filtering out noise from terrain variations at longer or shorter wavelengths.

@@

, where the resultant wavelet coefficient $C(s,x,y)$ provides a measure of how well the wavelet $\psi$ matches the data $z$ at each grid point [@Torrence1998]. 
When $s$ is large, $\psi$ is spread out, capturing long-wavelength features of $z$; when $s$ is small, $\psi$ becomes more localized, making it sensitive 
to fine-scale features of $z$. In this implementation, we use the 2D Ricker or Marr wavelet (i.e., Mexican hat wavelet) function to define $\psi$:

@@

The Mexican hat wavelet as defined above is equal to the second derivative of a Gaussian envelope, with its Fourier wavelength ($\lambda$) 
dependent on the chosen wavelet scale ($s$) and the grid spacing ($\Delta$) of the input DTM raster. The wavelet function $\psi$ is scaled according to the 
wavelet scale parameter $s$ and the grid spacing $\Delta$, ensuring that the resultant wavelet coefficient $C$ signifies 
concave and convex landforms corresponding to the wavelet scale $s$. Users can define the value of $\lambda$ in meters as the 
targeted spatial scale for landform roughness analysis, and the **pycwtmexhat.py** module will automatically compute the wavelet 
scale $s$ based on the grid spacing ($\Delta$) of the input raster file (Figure 1)\autoref{fig:1}.

We note that the equations for $C$ and $\psi$ presented here are mathematical approaches adapted in later publications 
[@LaHusen2020; @Underwood2022; @Herzig2023] and ongoing works on landslide mapping and age dating studies [@Booth2023; @Lai2023; @Ozioko2023]. 
There are minor differences in the proportionality constant used to define $\psi$ and the conventions used to present the magnitude of the wavelet 
coefficients $C$, compared to earlier similar research by @Booth2009 and @Booth2017 (original MATLAB codes 
available from [Booth's personal website](https://web.pdx.edu/~boothad/tools.html)). These differences in methatical approach will by definition 
result in $C$ values in different units and order of magnitude (e.g., $10^{-3}$ to $10^{-4}$ [m$^{-2}$] in @Booth2017 
and prior studies; $10^{-2}$ to $10^{-3}$ [m$^{-1}$] in @LaHusen2020 thereafter). Despite this discrepancy, the $C$ values 
yielded from these two approaches are linearly scaled and interconvertible, and they both reflect identical spatiotemporal patterns 
of topographic complexity (i.e., surface roughness).

@@

In this module, we adapt the variogram method to estimate the local fractal dimension within a moving window centered at each cell of 
the DTM [@Taud2005; @PardoIguzquiza2020; @PardoIguzquiza2022a]. This approach simplifies the problem to estimating the fractal dimension 
of one-dimensional topographic profiles [@Dubuc1989] within a two-dimensional moving window. For a one-dimensional profile of length $L$, 
the variogram $\gamma_1(k)$ can be estimated at the $K$ lag distances ($K$ lag distances ($k = 1, \ldots, K$) by:

@@

The **pyrugosity.py** module in **pyTopoComplexity**  measures the Rugosity Index of the land surface, which is widely used to assess structural 
complexity of topography and has been applied in classifying seafloor types by marine geologists and geomorphologists, understanding small-scale 
hydrodynamics by oceanographers, and studying available habitats in the landscape by ecologists and coral biologists [@Lundblad2006; @Wilson2007].

@@

This module adapts the Triangulated Irregular Networks method from @Jenness2004 to approximate the contoured area as the sum of 
eight truncated-triangle areas. These triangles connect the central grid point, four corner grid points, and four grid points at the middle points of the 
surrounding edges within the moving window. If no local slope correction is applied, the planimetric area is considered to be the horizontal 
planar area of the moving window, as described in @Jenness2004. Another approach considers slope correction where to the planimetric area is 
projected onto a plane of the local gradient [@DuPreez2015].

By definition, the Rugosity Index is as a minimum value of one (completely flat surface). Typical values of the conventional rugosity 
index (without slope correction) range from one to three although larger values are possible in very steep terrains. The slope-corrected rugosity 
index, also called arc-chord ratio (ACR) Rugosity Index, could provide a better representation of local surface complexity because it is not biased by slope (Figure 3)\autoref{fig:3}.

@@

The **pytpi.py** module in **pyTopoComplexity** calculates the Terrain Position Index (TPI) of the land surface. The TPI, also known as the 
Topographic Position Index in terrestrial studies [@Weiss2001], measures the relative topographic elevation of a point compared to those of its 
surrounding landforms. This metric highlights regions relatively higher or lower than their surroundings, which is useful for 
distinguishing landscape features such as hilltops, valleys, flat plains, and slopes. In oceanography, an equivalent metric is the 
Bathymetric Position Index (BPI), which applies the TPI algorithm to bathymetric data to evaluate seafloor complexity.

@@

In this module, the TPI is calculated for the central grid within a square moving window. Users can specify the size of the window (i.e., the 
number of grids along each edge) to evaluate topographic positions at various spatial scales. Positive TPI values indicate generally convex, 
elevated features, while negative values represent concave depressions. Values close to zero denote relatively planar areas. The **pytpi.py** 
module also returns the absolute values of the TPI (Figure 4\autoref{fig:4}). This absolute TPI raster is suggested to be effective for 
quick assessments in mapping and relative dating of deep-seated landslide deposits using high-resolution DTM data [@Lai2023].

@@

## Forward simulation of landscape smoothing through nonlinear hillslope diffusion processes

@@

, where \(\mathbf{q}_s\) represents the sediment flux at the surface. The sediment flux is further defined by a nonlinear flux law [@Roering1999] that is 
approximated using a Taylor series expansion [@Ganti2012]:

@@

Here, $\mathbf{S} = -\nabla z$ represents the downslope topographic gradient, and $S$ is its magnitude. $S_c$ is the critical slope representing the asymptotic 
maximum hillslope gradient. The parameter $K$ is a diffusion-like transport coefficient with dimensions of length squared per time. 
$N$ denotes the number of terms in the Taylor expansion, while $i$ specifies the number of additional terms included. If $N = 0$, the expression simplifies to 
linear diffusion [@Culling1963]. By default, $N$ is set to $2$, that gives the behavior described in @Ganti2012.

This notebook provides a comprehensive workflow that guides users through setting up, importing raster files, and running simulations. Since 
**Landlab** primarily handles DTM data in ESRI ASCII format, this notebook includes utility functions for converting raster files between GeoTIFF 
and ESRI ASCII formats. Users are required to specify the values for $S_c$, $K$, the length of each time step in years, and the final time to 
stop the simulation. The example included in the notebook uses lidar DTM data from the 2014 Oso Landslide [@WashingtonGeologicalSurvey2023], 
with parameter values provided in @Booth2017, in an attempt to reproduce the simulation results presented in that study (Figure 5\autoref{fig:5}).

@@

The development of **pyTopoComplexity** is part of a collaborative effort within the [Landslide subteam](https://cascadiacopeshub.org/team-1-landslides/) of the 
[Cascadia Coastlines and Peoples Hazards Research Hub (Cascadia CoPes Hub)](https://cascadiacopeshub.org), funded by the National Science Foundation Award ([#2103713](https://www.nsf.gov/awardsearch/showAward?AWD_ID=2103713&HistoricalAwards=false)). 
We also acknowledge earlier funding by the National Science Foundation to Booth ([#2000188](https://www.nsf.gov/awardsearch/showAward?AWD_ID=2000188&HistoricalAwards=false)) 
and Duvall ([#1953710](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1953710&HistoricalAwards=false)). We extend special thanks to Dr. Eulogio Pardo-Igúzquiza, 
who generously shared his Fortran code for fractal dimension analysis used in his work [@PardoIguzquiza2022b], which  inspired the development of the **pyfracd.py** module.
